Many claims of group polarization are plausibly false due to inappropriate statistics

### 2022-12-07 GroupPolarizationStatmod/paper/ms.tex
updated abstract and started updating figures to add sigma-finding diagram

\begin{abstract}
  \noindent
  Group polarization is the observed tendency of
  like-minded groups to arrive at a more extreme consensus opinion than
  the average opinion after discussing some topic of interest.
  Clearly it is important to understand such a phenomenon to understand broader
  problematic trends towards increased polariztion across society as a whole.
  Unfortunately, recent analyses have demonstrated that this measurement and
  analysis approach can result in false positive detections group polarization
  from failing to account for floor/ceiling effects induced when using ordinal scales to
  measure continuously-distributed latent psychological variables. 
  Here we demonstrate that many high profile detections
  of group polarization are plausibly false
  due to these floor/ceiling effects using generative modeling seeded by 
  reported group polarization values and parameters. After we identify which
  reported detections of group polarization are plausibly false detections, we further
  investigate whether the latent constant mean opinion and pre- and post-discussion
  opinion variances can fool the statistical tests used in each study to 
  generate false positives. Finally, we evaluate whether a more appropriate
  Bayesian cumulative normal statistical model can more accurately distinguish
  group polarization from simple consensus. Group polarization continues to 
  pose a problem for psychological theory with an ever-expanding set of potential theoretical
  explanations. A search for explanations is pointless, however, without sound
  evidence. We believe this work will provide a sturdier analytical foundation for 
  future group-polarization research.
\end{abstract}
\clearpage

@@

\begin{figure}
  \caption{Group polarization is experimentally induced by first gather
  participant responses to an opinion prompt and grouping together like-minded
  participants (A). Group polarization is theorized to occur when individuals
  are put together with like-minded individuals discuss the prompt (B). Group 
  polarization is detected by again polling all participants' opinions (C) and comparing
  these to the initial opinions.}
  \label{fig:groupPolarizationSchematic}
  \centering
  \includegraphics[width=\textwidth]{Figures/GPExperimentSketch.pdf}
\end{figure}

@@

\section{Model and Data}

Our model represnts the generative process of finding different forms of group consensus
in group polarization experiments, and the process of measuring that consensus using
an ordinal measurement scale, e.g.\ a Likert-type scale. This model considers two
relevant forms of consensus: \emph{simple consensus} and \emph{group polarization}.
Either process is \emph{consensus}, if we define consensus as a reduction
in opinion variance over time.  
Group polarization experiments are designed to result group consensus
since they group together like-minded individuals to see if they become more
extreme after discussion~\cite{Turner2018}.

@@

Our model represents continuous latent opinions and their 
measurement on an ordinal scale. \emph{Latent opinions} are 
unobservable, psychological variables that are 
somehow represented and stored in the brain-body, and can be reproduced with
some accuracy when a person is prompted to give their opinion. 
\emph{Ordinal measurements} of 
of latent opinions are discrete-valued representations of latent opinions 
in some physical form, such as a mark on a sheet of paper or radio button on a web form, that
indicate the polarity and strength of a person's opinion opinion in terms of
one of a number of opinion bin values. All group polarization experiments must somehow
measure opinions, and all but a few use ordinal measurements (those few exceptions include
using bets made during blackjack as a proxy for opinions~\cite{Blascovich1973,Blascovich1975,Blascovich1976}
and using mock jury punitive rewards as a proxy for opinions~\cite{Schkade2000}).

% For example, on a 7-point ordinal scale -3 could indcate 
% ``strongly disagree'', +3 could be ``strongly agree'', 0 would be psychologically 
% neutral, and intermediate values are less extreme opinions of either valence, 
% such as was used by \citeA{Moscovici1969}. Or, one might

>>>>

\begin{abstract}
  \noindent
  Psychology is beset by three crises: the reproducibility crisis, the
  generalizability crisis, and the theory crisis. New verbal theories for various
  phenomena are constantly being developed orthogonal to other theories, all based on
  mostly unreproducible analyses; even if a positive result is in some sense ``true''
  it may not account for all the variance present in a set of observations---studies
  thus fail to generalize because their results do not account for this variance.
  There is another related problem: a phenomenon may be ``reproducible'' only
  because studies of the phenomenon systematically misuse statistics in such a way
  that induces false positives for each succeeding ``replication.'' Here we show that
  replicated detections of group polarization are often plausibly false due to
  treating ordinal measurements of opinions, e.g., on a Likert-type scale, as drawn
  directly from an underlying continuous latent opinion distribution, which masks
  extreme latent opinions. Our intensive review of sixty experimental conditions
  across ten foundational group polarization papers revealed that 95\% of the
  conditions generated plausibly false positive results. To show this, we developed a
  novel generative model that provides two levels of evidence that an historical
  published empirical finding of group polarization is plausibly false.  Our model
  can be used to evaluate additional studies in group polarization and beyond, even
  if the reported data and significant metrics are incompletely reported, and the
  original raw data has been lost to time.  This may be just the tip of the
  plausibly-false-positives-in-psychology iceberg, both for group polarization
  studies and for the legion psychological results that model ordinal observations of
  opinions, beliefs, emotions, or any other latent psychological variable that is
  theoretically continuous, but measured using an ordinal scale.  
\end{abstract}
\clearpage

@@

\begin{figure}
  \caption{Group polarization is experimentally induced by first gathering
  participant responses to an opinion prompt (A). Then groups of like-minded 
  individuals discuss the prompt (B). Group 
  polarization occurs if the mean post-discussion opinion is more extreme 
  than the mean pre-discussion opinion (C).}
  \label{fig:groupPolarizationSchematic}
  \centering
  \includegraphics[width=\textwidth]{Figures/GPExperimentSketch.pdf}
\end{figure}

@@

\section{Model}

Our model represnts the generative process of finding different forms of group consensus
in group polarization experiments, and the process of measuring that consensus using
an ordinal measurement scale, e.g.\ a Likert-type scale. This model operationalizes two
relevant forms of consensus: \emph{simple consensus} and \emph{group polarization}.
\mt{Need to explain in Intro why consensus should occur along with group
polarization} 
% Group polarization experiments are designed to result group consensus
% since they group together like-minded individuals to see if they become more
% extreme after discussion~\cite{Turner2018}.

@@
